Diversity dynamics in Nymphalidae butterflies: Effect of phylogenetic uncertainty on diversification rate shift estimates

### 2014-11-05 MS_revised.tex
ref 49

The diffuse cospeciation hypothesis predicts almost identical ages of
insects and their hostplants, while the ``resource abundance-dependent
diversity'' and the ``escape-and-radiate'' hypotheses state that insects
diversify after their hostplants {[}2--4{]}. Wheat et al. (2007)
{[}47{]} found strong evidence for a model of speciation congruent with
Ehrlich and Raven's hypothesis in Pieridae butterflies due to, in
addition to the identification of a key innovation, a burst of
diversification in glucosinolate-feeding taxa shortly afterwards (with a
lag of circa 10 MY). According to a recent dated phylogeny of
the Angiosperms {[}49{]}, the family Solanaceae split from its sister
group about 59 (49--68) MYA and diversification started (crown group
age) around 37 (29--47) MYA. Wahlberg et al. (2009) {[}13{]} give the
ages for origin and diversification for Ithomiini at 45 (39--53) and 37
(32--43) MYA, respectively. Thus, current evidence shows that Solanaceae
and Ithomiini might have diversified around the same time, during the
Late Eocene and Oligocene, and this would be congruent with the diffuse
cospeciation hypothesis.

\subsubsection*{Danaini}
Our MultiMEDUSA approach showed a significant slowdown in net
diversification rate in the subtribe Danaina of the Danini. Both Danaina
and the sister clade Euploeina feed mainly on Apocynaceae and thus a
hostplant shift should not be responsible for the observed slowdown of
diversification in the Danaina. As expected, our BiSSE analysis of
Apocynaceae feeders shows that there is no effect of feeding on this
plant family on the net diversification rates of Nymphalidae lineages.
Many of the Danaina are large, strong fliers, highly migratory and
involved in mimicry rings. Among them is for example the monarch
(\emph{Danaus plexippus}), probably the most well known of all migratory
butterflies. The causes for a lower net diversification rate in the
Danaina remains to be investigated, but their great dispersal power
might be involved in preventing allopatric speciation. It has been found
in highly vagile species in the nymphalid genus \emph{Vanessa} that
dispersal has homogenized populations due to gene flow, as old and
vagile species seem to be genetically homogeneous while younger
widespread species show higher genetic differentiation in their
populations {[}50{]}.

>>>>

The diffuse cospeciation hypothesis predicts almost identical ages of
insects and their hostplants, while the ``resource abundance-dependent
diversity'' and the ``escape-and-radiate'' hypotheses state that insects
diversify after their hostplants {[}2--4{]}. Wheat et al. (2007)
{[}47{]} found strong evidence for a model of speciation congruent with
Ehrlich and Raven's hypothesis in Pieridae butterflies due to, in
addition to the identification of a key innovation, a burst of
diversification in glucosinolate-feeding taxa shortly afterwards (with a
lag of circa 10 MY). According to a recent dated phylogeny of
the Angiosperms {[}50{]}, the family Solanaceae split from its sister
group about 59 (49--68) MYA and diversification started (crown group
age) around 37 (29--47) MYA. Wahlberg et al. (2009) {[}13{]} give the
ages for origin and diversification for Ithomiini at 45 (39--53) and 37
(32--43) MYA, respectively. Thus, current evidence shows that Solanaceae
and Ithomiini might have diversified around the same time, during the
Late Eocene and Oligocene, and this would be congruent with the diffuse
cospeciation hypothesis.

\subsubsection*{Danaini}
Our MultiMEDUSA approach showed a significant slowdown in net
diversification rate in the subtribe Danaina of the Danini. Both Danaina
and the sister clade Euploeina feed mainly on Apocynaceae and thus a
hostplant shift should not be responsible for the observed slowdown of
diversification in the Danaina. As expected, our BiSSE analysis of
Apocynaceae feeders shows that there is no effect of feeding on this
plant family on the net diversification rates of Nymphalidae lineages.
Many of the Danaina are large, strong fliers, highly migratory and
involved in mimicry rings. Among them is for example the monarch
(\emph{Danaus plexippus}), probably the most well known of all migratory
butterflies. The causes for a lower net diversification rate in the
Danaina remains to be investigated, but their great dispersal power
might be involved in preventing allopatric speciation. It has been found
in highly vagile species in the nymphalid genus \emph{Vanessa} that
dispersal has homogenized populations due to gene flow, as old and
vagile species seem to be genetically homogeneous while younger
widespread species show higher genetic differentiation in their
populations {[}51{]}.
